An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-06-22 cal_demo/ms.tex
Minor fixes when rereading text

We conclude this section by connecting our calibration expression to that found in a visibility 
framework. In the limit of a single bright calibrating source at center, we can greatly 
simplify equations~\ref{eq:Cna} and~\ref{eq:cal_solution}. We will assume the radiation patterns are 
normalized such that $\mathcal{W}(0)=1$. We can further drop the exponential phase terms 
because $\spix$ is perpendicular to our planar antenna array. We then absorb the true gains into the true visibilities in 
equation~\ref{eq:Cna} to express as a sum of measured, uncalibrated, visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_b \frac{1}{g^{*(n)}_b} V_{ab}
\end{equation}

We next plug this expression into equation~\ref{eq:cal_solution} to find our simplified 
calibration solution for a single bright point source. Because our sky is a single bright point 
source, the model visibilities are simply the flux of the source, $S_{\mathrm{src}}$.
\begin{equation}
g^{(n+1)}_a \rightarrow \frac{\sum_b  V_{ab}/g^{*(n)}_b}{\Nant S_{\mathrm{src}}}
\end{equation}
This is simply a gain-weighted sum of the measured visibilities over the flux of the source, 
which is indeed the limiting result from a visibility approach, for example seen in \citealt{mit08}. 
The ability to recover the equivalent expression despite not actually forming the visibilities is a 
result of the fact that only sums over visibilities come into the visibility-based solution, as was described in 
\citealt{mor11}. We have confirmed the limiting case equivalence here, and will explore the 
more general case in more detail in the following sections.

@@

For our unknown gains, we create a set of random complex numbers where the amplitude is 
drawn from approximately gaussian distribution centered around 1 with width 0.25, and 
completely random phase. These are our ``true gains", and we apply them to the frequency
domain simulated antenna electric field measurements as in equation~\ref{eq:apply_gain}. Our analysis is 
blind to these values until the end of the process to check accuracy. The gain estimates are 
initialized with unity, $g^{(0)}_a=1$.

@@

Images created at the beginning of calibration and at the end are shown in 
Fig.~\ref{fig:sim_images}. Each image is obtained over 400 time sample integration, corresponding to 
all snapshot images created with a given set of gain estimates. The left panel shows the image 
produced with our initialized unity gains. Because the phases are completely random, the 
image is essentially noise with the primary beam evident. After 20 calibration iterations, the image is far 
more clear, shown in the middle panel. 
The ten sources are visible, along with rumble throughout the image. By comparing with
a perfectly calibrated image (right panel), we see that the rumble and the shapes of the
point sources are matched between the two images. We therefore conclude that the 
artifacts visible are dominated by the point spread function, rather than calibration errors.

\begin{figure*}
\begin{center}
\includegraphics[width=.3\linewidth]{fig5a.pdf}
\includegraphics[width=.3\linewidth]{fig5b.pdf}
\includegraphics[width=.3\linewidth]{fig5c.pdf}
\caption{Images formed during simulated calibration. \emph{Left:} An image generated from a 
single frequency channel and 400 time sample integration after the gain estimates are randomized. As 
expected with random phases, the image is completely noisy, with the shape of the primary 
beam evident. \emph{Middle:} An image formed after calibration, again with a single frequency 
channel and 400 time sample integration. Now the simulated point sources are visible in the image. 
\emph{Right:} An image generated from the same simulated data as the middle image, but 
with perfect calibration correction. Many of the features seen in the middle image are also
seen here, confirming that they are point spread function artifacts. The ten simulated (and modeled) point sources are highlighted with red circles.
\label{fig:sim_images}
}
\end{center}
\end{figure*}

@@

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The 
left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent 
near the center of the image, and Cas~A is also clearly visible in the upper right. The middle 
panel shows the image produced after calibration with identical integration time and bandwidth. 
The sidelobes throughout the image are significantly suppressed, and the galactic plane is 
much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature 
just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux 
from the GSM. 
For reference we show the GSM in the right panel of 
Fig.~\ref{fig:data_images}, convolved by the LWA point spread function and weighted
by  two factors of the primary beam to match the holographic frame of the data images.

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{fig9a.pdf}
\includegraphics[width=0.3\linewidth]{fig9b.pdf}
\includegraphics[width=0.3\linewidth]{fig9c.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images 
were produced with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows 
significant reduction in rumble throughout, while retaining the prominent Cyg~A and 
Cas~A sources. The galactic plane is also substantially more evident after calibration. For 
reference, the GSM is shown in the 
right panel convolved with the LWA point spread function and weighted by two factors
of the primary beam to place it in the same holographic frame as the data images. While much of the GSM is visible in the calibrated image, we 
note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

@@

The work here will serve as a foundation for further development. We have shown that the 
EPICal algorithm produces reliable calibration solutions, and have identified several aspects to 
increase the scope. With next generation instruments in the planning and development stages, 
EPICal is poised to greatly contribute to their success. The software is integrated into the EPIC 
package and freely available to use in simulations or offline processing of data. Work is 
underway to port the code to real-time GPU systems for deployment.

>>>>

We conclude this section by connecting our calibration expression to that found in a visibility 
framework. In the limit of a single bright calibrating source at center, we can greatly 
simplify equations~\ref{eq:Cna} and~\ref{eq:cal_solution}. We will assume the radiation patterns are 
normalized such that $\mathcal{W}(0)=1$. We can further drop the exponential phase terms 
because $\spix$ is perpendicular to our planar antenna array. We then absorb the true gains into the true visibilities in 
equation~\ref{eq:Cna} to express as a sum of measured, uncalibrated, visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_{b\ne a} \frac{1}{g^{*(n)}_b} V_{ab}
\end{equation}

We next plug this expression into equation~\ref{eq:cal_solution} to find our simplified 
calibration solution for a single bright point source. Because our sky is a single bright point 
source, the model visibilities are simply the flux of the source, $S_{\mathrm{src}}$.
\begin{equation}
g^{(n+1)}_a \rightarrow \frac{\sum_{b\ne a}  V_{ab}/g^{*(n)}_b}{\Nant S_{\mathrm{src}}}
\end{equation}
This is simply a gain-weighted sum of the measured visibilities over the flux of the source, 
which is indeed the limiting result from a visibility approach, for example seen in \citealt{mit08}. 
The ability to recover the equivalent expression despite not actually forming the visibilities is a 
result of the fact that only sums over visibilities come into the visibility-based solution, as was described in 
\citealt{mor11}. We have confirmed the limiting case equivalence here, and will explore the 
more general case in more detail in the following sections.

@@

For our unknown gains, we create a set of random complex numbers where the amplitude is 
drawn from an approximately gaussian distribution centered around 1 with width 0.25, and 
completely random phase. These are our ``true gains", and we apply them to the frequency
domain simulated antenna electric field measurements as in equation~\ref{eq:apply_gain}. Our analysis is 
blind to these values until the end of the process to check accuracy. The gain estimates are 
initialized with unity, $g^{(0)}_a=1$.

@@

Images created at the beginning of calibration and at the end are shown in 
Fig.~\ref{fig:sim_images}. Each image is obtained over 400 time sample (10 ms) integration, corresponding to 
all snapshot images created with a given set of gain estimates. The left panel shows the image 
produced with our initialized unity gains. Because the phases are completely random, the 
image is essentially noise with the primary beam evident. After 20 calibration iterations, the image is far 
more clear, shown in the middle panel. 
The ten sources are visible, along with rumble throughout the image. By comparing with
a perfectly calibrated image (right panel), we see that the rumble and the shapes of the
point sources are matched between the two images. We therefore conclude that the 
artifacts visible are dominated by the point spread function, rather than calibration errors.

\begin{figure*}
\begin{center}
\includegraphics[width=.3\linewidth]{fig5a.pdf}
\includegraphics[width=.3\linewidth]{fig5b.pdf}
\includegraphics[width=.3\linewidth]{fig5c.pdf}
\caption{Images formed during simulated calibration. \emph{Left:} An image generated from a 
single frequency channel and 400 time sample (10 ms) integration after the gain estimates are randomized. As 
expected with random phases, the image is completely noisy, with the shape of the primary 
beam evident. \emph{Middle:} An image formed after calibration, again with a single frequency 
channel and 400 time sample integration. Now the simulated point sources are visible in the image. 
\emph{Right:} An image generated from the same simulated data as the middle image, but 
with perfect calibration correction. Many of the features seen in the middle image are also
seen here, confirming that they are point spread function artifacts. The ten simulated (and modeled) point sources are highlighted with red circles.
\label{fig:sim_images}
}
\end{center}
\end{figure*}

@@

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The 
left panel shows the uncalibrated image integrated over 51.2 ms, 83 kHz. Cyg~A is prominent 
near the center of the image, and Cas~A is also clearly visible in the upper right. The middle 
panel shows the image produced after calibration with identical integration time and bandwidth. 
The sidelobes throughout the image are significantly suppressed, and the galactic plane is 
much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature 
just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux 
from the GSM. 
For reference we show the GSM in the right panel of 
Fig.~\ref{fig:data_images}, convolved by the LWA point spread function and weighted
by  two factors of the primary beam to match the holographic frame of the data images.

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{fig9a.pdf}
\includegraphics[width=0.3\linewidth]{fig9b.pdf}
\includegraphics[width=0.3\linewidth]{fig9c.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images 
were produced with 83 kHz bandwidth and 51.2 ms integration. The calibrated image shows 
significant reduction in rumble throughout, while retaining the prominent Cyg~A and 
Cas~A sources. The galactic plane is also substantially more evident after calibration. For 
reference, the GSM is shown in the 
right panel convolved with the LWA point spread function and weighted by two factors
of the primary beam to place it in the same holographic frame as the data images. While much of the GSM is visible in the calibrated image, we 
note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

@@

The work here will serve as a foundation for further development. We have shown that the 
EPICal algorithm produces reliable calibration solutions, and have identified several aspects to 
increase the scope. With next generation instruments in the planning and development stages, 
EPICal is poised to greatly contribute to their success. The software is integrated into the EPIC 
package and freely available to use in simulations or post-processing of data. Work is 
underway to port the code to real-time GPU systems for deployment.
